---
title: "How to Calculate bete Coefficients"
author: "Belias Michael"
date: "4 April 2016"
output:
  pdf_document:
    highlight: pygments
    number_sections: yes
    toc: yes
    toc_depth: 3
  html_document:
    css: APAStyle.css
    fig_caption: yes
    highlight: espresso
    theme: null
  word_document:
    reference: APA.docx
fontsize: 12pt
csl: Harvard2.csl
---



\newpage 


# Introduction


# Beta Calculation


The linear model is writen as __$Y= X\beta + \epsilon$__ where Y denotes the vector of responces, $\beta$ is the vector of coefficients, **X** is the corresponding design matrix  and $\epsilon$ is the vector of random errors. 

The $\beta$ is given by the :

$\hat \beta =(X'X)^{-1}X'Y$ 

and 

$$Var(\hat \beta)=\sigma^2(X'X)^{-1}$$ aka.




Let's create a Data-Set with $\epsilon$ ~ N(0 , 4)

```{r}
set.seed(1152)
n <- 1000     #nb of observations
b <- vector()       #intercept
x <- matrix(nrow = n,ncol = 100)
x.b<- matrix(nrow = n,ncol = 100)
for(i  in 2:100){
  b[i] = runif(1, min = -10, max = 10)
  x[,i] = rnorm(n = n,mean = runif(1, min = -10, max = 10) , 
                sd= runif(1, min = 0, max = 4))
}


x[,1]= rep(1)
b[1] = 7.3

epsilon <- rnorm(n, mean=0, sd=sqrt(4))

y <- b*x + epsilon

df= data.frame(y=y,x=x[,-1])
fit <- lm(y ~ x.2 +x.3 +x.4 , df)
```

Let's use the explicit formulas.


$$\hat Var (\hat \beta)=\hat \sigma^2(X'X)^{-1}$$
```{r}

X <- cbind(1, x1,x2)
n<-dim(X)[2]
betaHat <- solve(t(X) %*% X) %*% t(X) %*% y
var_betaHat <- anova(fit)[[n]][n] * solve(t(X) %*% X)
```


Let's compare the coefficients: 
```{r}
library(knitr)
a=cbind(fit$coef,c(betaHat[1], betaHat[2],betaHat[3]))
colnames(a)<- c("Model-Estimated","Manually-Estimated")
kable(a,caption="Comparison of the Betas")
```

Let's compare the Standard Errors:

```{r}
 a=cbind(summary(fit)$coefficients[, 2],sqrt(diag(var_betaHat)))
colnames(a)<- c("Model-Estimated","Manually-Estimated")
kable(a,caption="Comparison of the Betas")
```

